---
title: 'R Project: Fernando Notes'
author: "Fernando Szeliga"
date: "2023-11-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
remotes::install_github("nrennie/LondonMarathon")
data(winners, package = "LondonMarathon")
data(london_marathon, package = "LondonMarathon")

library(tidyverse)

```
## Research Questions 

- Nationalities
- Time to complete

##  Steps/Analyses

# Join the data

```{r}

joint_marathondata <- dplyr::full_join(london_marathon, winners, by="Year")

```

# Time Transformation

```{r}

# The time column is in fraction of days,so this was converted to seconds
time_in_seconds <- as.numeric(joint_marathondata$Time) * 24 * 60 * 60

# Formatting the time components
joint_marathondata$Time_Converted <- format(as.POSIXct(time_in_seconds,origin = "1970-01-01", tz = "UTC"), format = "%H:%M:%S")

```

# Continents (new column)

```{r}

winners_countries <- unique(joint_marathondata$Nationality)


continent_vector <- sapply(joint_marathondata$Nationality, function(x)
{
  
  if (x== "Norway" | x=="Ireland" |x== "Denmark" |x== "Poland" |x== "Sweden" | x=="Portugal" |x== "France" |x== "Germany" |x== "Belgium" |x== "Switzerland" |x== "Spain" |x== "Italy" | x=="Netherlands")
  {
    return("Rest of Europe")
  }
    else if (x== "United Kingdom")
  {
    return("United Kingdom")
  }
  else if (x== "United States" |x== "Canada" |x== "Mexico")
  {
    return("North America")
  }
  else if (x== "Japan" |x== "Soviet Union" |x== "China")
  {
    return("Asia")
  }
  else if (x== "Kenya" |x== "Morocco" |x== "Ethiopia")
  {
    return("Africa")
  }
  else (x == "Australia")
  {
    return("Oceania")
  }
  
  continent_vector <- continent_vector
  
})


joint_marathondata$Winner_Continent <- continent_vector

```

# Continent Plots

```{r}

ggplot(data = joint_marathondata) +
  geom_bar(mapping = aes(x = fct_infreq(Winner_Continent)), fill= "blue") + labs(x= "Geographic Area", y= "Winners", title =  "Winner's Geographic Distribution")

ggplot(data = joint_marathondata) +
  geom_bar(mapping = aes(x = Winner_Continent, fill=Category), position = "fill") + labs(x= "Geographic Area", y= "Density", title =  "Winner's Geographic Distribution by Competition") + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

```

# Countries


NOTE that there is a tie in 1981 between a participant from Norway and a participant from UK. 

```{r}

ggplot(data = joint_marathondata) +
  geom_bar(mapping = aes(x = fct_infreq(Nationality)), fill= "blue") + labs(x= "Nationality", y= "Winners", title =  "Winner's Nationality") + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

ggplot(data = joint_marathondata) +
  geom_bar(mapping = aes(x = Nationality, fill=Category), position = "fill") + labs(x= "Nationality", y= "Density", title =  "Winner's Nationality by Competition") + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

```
# Time comparison. Category: Men

```{r}

winners |> filter(Category=="Men") |> group_by(Nationality) |> summarise(finish_time= mean(Time)) -> men_category


# The time column is in fraction of days,so this was converted to seconds
time_men <- as.numeric(men_category$finish_time) * 24 * 60 * 60

# Formatting the time components
men_category$time_conversion <- format(as.POSIXct(time_men,origin = "1970-01-01", tz = "UTC"), format = "%H:%M:%S")

ggplot(men_category, aes(x = reorder(Nationality,finish_time, decreasing=TRUE), y = time_conversion)) +
  geom_point() + geom_segment(aes(xend = fct_reorder(Nationality,finish_time)), yend = 0, col = "blue") +
  coord_flip() + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) + xlab("Nationality")+ ylab("Finish Time (Mean)")

```
# Time comparison. Category: Women

```{r}

winners |> filter(Category=="Women") |> group_by(Nationality) |> summarise(finish_time= mean(Time)) -> women_category


# The time column is in fraction of days,so this was converted to seconds
time_women <- as.numeric(women_category$finish_time) * 24 * 60 * 60

# Formatting the time components
women_category$time_conversion <- format(as.POSIXct(time_women,origin = "1970-01-01", tz = "UTC"), format = "%H:%M:%S")

ggplot(women_category, aes(x = reorder(Nationality,finish_time, decreasing=TRUE), y = time_conversion)) +
  geom_point() + geom_segment(aes(xend = fct_reorder(Nationality,finish_time)), yend = 0, col = "blue") +
  coord_flip() + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) + xlab("Nationality")+ ylab("Finish Time (Mean)")

```


# Time comparison. Category: Wheelchair Men


```{r}

winners |> filter(Category=="Wheelchair Men") |> group_by(Nationality) |> summarise(finish_time= mean(Time)) -> wmen_category


# The time column is in fraction of days,so this was converted to seconds
time_wmen <- as.numeric(wmen_category$finish_time) * 24 * 60 * 60

# Formatting the time components
wmen_category$time_conversion <- format(as.POSIXct(time_wmen,origin = "1970-01-01", tz = "UTC"), format = "%H:%M:%S")

ggplot(wmen_category, aes(x = reorder(Nationality,finish_time, decreasing=TRUE), y = time_conversion)) +
  geom_point() + geom_segment(aes(xend = fct_reorder(Nationality,finish_time)), yend = 0, col = "blue") +
  coord_flip() + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) + xlab("Nationality")+ ylab("Finish Time (Mean)")

```

# Time comparison. Category: Wheelchair Women

```{r}

winners |> filter(Category=="Wheelchair Women") |> group_by(Nationality) |> summarise(finish_time= mean(Time)) -> wwomen_category


# The time column is in fraction of days,so this was converted to seconds
time_wwomen <- as.numeric(wwomen_category$finish_time) * 24 * 60 * 60

# Formatting the time components
wwomen_category$time_conversion <- format(as.POSIXct(time_wwomen,origin = "1970-01-01", tz = "UTC"), format = "%H:%M:%S")

ggplot(wwomen_category, aes(x = reorder(Nationality,finish_time, decreasing=TRUE), y = time_conversion)) +
  geom_point() + geom_segment(aes(xend = fct_reorder(Nationality,finish_time)), yend = 0, col = "blue") +
  coord_flip() + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) + xlab("Nationality")+ ylab("Finish Time (Mean)")

```

