---
title: "R_Project_Analysis"
author: "Akshay Suresh Varma"
date: "2023-12-4"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Loading Packages
```{r}
library(tidyverse)
library(GGally)
library(MASS)
library(mosaicData)

```

```{r}
remotes::install_github("nrennie/LondonMarathon")
data(winners, package = "LondonMarathon")
data(london_marathon, package = "LondonMarathon")

```

```{r}

winners_data <- winners
winners_data

marathon_data <- london_marathon
marathon_data
```

```{r}
library(ggplot2)

ggplot(winners_data, aes(x = Year, y = ..count.., fill = Nationality)) +
  geom_bar(position = "stack") +
  labs(title = "Nationality of Winners Over the Years",
       x = "Year",
       y = "Count",
       fill = "Nationality") +
  facet_wrap(~Category, scales = "free_y")
```


```{r}
library(ggplot2)

ggplot(winners_data[winners_data$Category == "Men", ], aes(x = Year, fill = Nationality)) +
  geom_bar(position = "stack") +
  labs(title = "Nationality of Men Winners Over the Years",
       x = "Year",
       y = "Count",
       fill = "Nationality")

ggplot(winners_data[winners_data$Category == "Women", ], aes(x = Year, fill = Nationality)) +
  geom_bar(position = "stack") +
  labs(title = "Nationality of Women Winners Over the Years",
       x = "Year",
       y = "Count",
       fill = "Nationality")

ggplot(winners_data[winners_data$Category == "Wheelchair Men", ], aes(x = Year, fill = Nationality)) +
  geom_bar(position = "stack") +
  labs(title = "Nationality of Wheelchair Men Winners Over the Years",
       x = "Year",
       y = "Count",
       fill = "Nationality")

ggplot(winners_data[winners_data$Category == "Wheelchair Women", ], aes(x = Year, fill = Nationality)) +
  geom_bar(position = "stack") +
  labs(title = "Nationality of Wheelchair Women Winners Over the Years",
       x = "Year",
       y = "Count",
       fill = "Nationality")

```


Aspect 1: Winners' Nationalities
```{r}
# Load required libraries
library(dplyr)


# Count the number of wins by nationality
nationality_counts <- winners_data %>%
  group_by(Category, Nationality) %>%
  summarise(Wins = n()) %>%
  arrange(desc(Wins))

print("Winners' Nationalities:")
print(nationality_counts)

```



Aspect 2: Winning Times
```{r}

# Visualize winning times by category
library(ggplot2)

ggplot(winners_data, aes(x = Year, y = Time, color = Category)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Winning Times Over the Years",
       x = "Year",
       y = "Winning Time",
       color = "Category") +
  theme_minimal()

```



Aspect 3: Trends in Winning Times Over Years
```{r}

# Visualize trends in winning times over the years
ggplot(winners_data, aes(x = Year, y = Time, color = Category)) +
  geom_line() +
  labs(title = "Trends in Winning Times Over the Years",
       x = "Year",
       y = "Winning Time",
       color = "Category") +
  theme_minimal()
```


Aspect 4: Winning Athletes
```{r}

# Identify athletes with the most wins
top_athletes <- winners_data %>%
  group_by(Category, Athlete, Nationality) %>%
  summarise(Wins = n()) %>%
  arrange(desc(Wins))

print("Top Athletes with Most Wins:")
print(top_athletes)

```


Aspect 1: Winners' Nationalities
```{r}

library(ggplot2)

# Convert the 'Year' column to a factor to maintain the order
marathon_data$Year <- as.factor(marathon_data$Year)

# Plotting the trend of applicants over the years
ggplot(marathon_data, aes(x = Year, y = Applicants)) +
  geom_line() +
  labs(title = "Number of Applicants Over the Years",
       x = "Year",
       y = "Number of Applicants")

```


Bar plot
```{r}
# Bar plot of the number of applicants over the years
ggplot(marathon_data, aes(x = Year, y = Applicants)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(title = "Number of Applicants Over the Years",
       x = "Year",
       y = "Number of Applicants")


# Define the subset of years for better readability
subset_years <- c(1990, 1995, 2000, 2005, 2010, 2015, 2020)

# Bar plot of the number of applicants over the years with adjusted axis labels
ggplot(marathon_data, aes(x = factor(Year), y = Applicants)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(title = "Number of Applicants Over the Years",
       x = "Year",
       y = "Number of Applicants") +
  scale_x_discrete(labels = function(x) ifelse(as.numeric(x) %in% subset_years, x, ""))

```

```{r}


# Calculate acceptance rate
marathon_data$Acceptance_Rate <- (marathon_data$Accepted / marathon_data$Applicants) * 100

# Line plot of acceptance rate over the years
ggplot(marathon_data, aes(x = Year, y = Acceptance_Rate)) +
  geom_line(color = "blue") +
  geom_point(color = "blue") +
  labs(title = "Acceptance Rate Over the Years",
       x = "Year",
       y = "Acceptance Rate (%)")


```




1)Number of Applicants Over the Years:

Visualized using a line plot to show the trend.
Increasing trend indicates growing interest in the marathon.
```{r}
ggplot(marathon_data, aes(x = Year, y = Applicants)) +
  geom_line(color = "green") +
  geom_point(color = "green") +
  labs(title = "Number of Applicants Over the Years",
       x = "Year",
       y = "Number of Applicants")
```


2)Number of Finishers Over the Years:

Visualized using a line plot.
Indicates the growth in the number of participants who successfully completed the marathon.
```{r}
ggplot(marathon_data, aes(x = Year, y = Finishers)) +
  geom_line(color = "red") +
  geom_point(color = "red") +
  labs(title = "Number of Finishers Over the Years",
       x = "Year",
       y = "Number of Finishers")
```


3)Acceptance Rate Over the Years:

Calculated and visualized using a line plot.
Helps understand the competitiveness of the marathon.
```{r}
marathon_data$Acceptance_Rate <- (marathon_data$Accepted / marathon_data$Applicants) * 100

ggplot(marathon_data, aes(x = Year, y = Acceptance_Rate)) +
  geom_line(color = "blue") +
  geom_point(color = "blue") +
  labs(title = "Acceptance Rate Over the Years",
       x = "Year",
       y = "Acceptance Rate (%)")
```

4)Nationality of Winners Over the Years (by Category):

Visualized using a stacked bar plot with facets for each category.
Provides insights into the diversity of winners by nationality.
```{r}
ggplot(winners_data, aes(x = Year, y = ..count.., fill = Nationality)) +
  geom_bar(position = "stack") +
  labs(title = "Nationality of Winners Over the Years",
       x = "Year",
       y = "Count",
       fill = "Nationality") +
  facet_wrap(~Category, scales = "free_y")
```

5)Time Distribution of Marathon Winners (by Category):

Box plot to show the distribution of winning times.
Helps identify trends in performance over the years.
```{r}
ggplot(winners_data, aes(x = Category, y = Time, fill = Category)) +
  geom_boxplot() +
  labs(title = "Time Distribution of Marathon Winners",
       x = "Category",
       y = "Time (hours)",
       fill = "Category")
```


```{r}
ggplot(winners_data, aes(x = Year, y = ..count.., fill = Nationality)) +
  geom_bar(position = "stack") +
  labs(title = "Nationality of Winners Over the Years",
       x = "Year",
       y = "Count",
       fill = "Nationality") +
  facet_wrap(~Category, scales = "free_y")
```


```{r}
```


```{r}

```


```{r}

```

```{r}

```
