---
title: "Sam's Piece"
author: "Sam Alexander White"
date: "2023-12-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Reading Files and importing libraries
```{r}
data(winners, package = "LondonMarathon")
data(london_marathon, package = "LondonMarathon")
library(dplyr)
library(ggplot2)
library(tidyr)
```

Adding Probability Columns (Code From 1st Meeting)
```{r}
# Probability of applicants being accepted
prob_accepted <- london_marathon %>%
  group_by(Year) %>%
  summarize(prob_accepted = ifelse(sum(Applicants, na.rm = TRUE) == 0, 0, 
                          sum(Accepted,na.rm=TRUE)/sum(Applicants,na.rm=TRUE)))

# Probability of an accepted athlete starting the race
prob_starting <- london_marathon %>%
  group_by(Year) %>%
  summarize(prob_starting = ifelse(sum(Accepted, na.rm = TRUE) == 0, 0, 
                     sum(Starters, na.rm = TRUE) / sum(Accepted, na.rm = TRUE)))

# Calculate the probability of a starting athlete finishing the race
prob_finishing <- london_marathon %>%
  group_by(Year) %>%
  summarize(prob_finishing = ifelse(sum(Starters, na.rm = TRUE) == 0, 0, 
                    sum(Finishers, na.rm = TRUE) / sum(Starters, na.rm = TRUE)))

# Add the calculated probabilities to the original data set
london_marathon <- london_marathon %>%
  left_join(prob_accepted, by = "Year") %>%
  left_join(prob_starting, by = "Year") %>%
  left_join(prob_finishing, by = "Year")

# Add a column for Winning Probability (using total applicants)
london_marathon <- london_marathon %>%
  group_by(Year) %>%
  mutate(Winning_Prob_Total = ifelse(sum(Applicants,
                              na.rm=TRUE)==0,0,1/sum(Applicants, na.rm = TRUE)))

# Add a column for Winning Probability (using accepted applicants)
london_marathon <- london_marathon %>%
  group_by(Year) %>%
  mutate(Winning_Prob_Accepted = ifelse(sum(Accepted, 
                        na.rm = TRUE) == 0, 0, 1 / sum(Accepted, na.rm = TRUE)))

# Print the updated data set
print(london_marathon)
```

Mean Times By Cateogry (Using Akshay's code I think)
```{r}
# Convert 'Time' to numeric (if it's not already)
winners$Time <- as.numeric(winners$Time)

# Aggregate the data to find the average 'Time' for each 'Category'
average_time_by_category <- aggregate(Time ~ Category, data = winners, FUN = mean)

# Display the result
print(average_time_by_category)
```

New Data (Prepared For Second Meeting)
Sam's Q - Are Men's times faster than women's? by how much?
```{r}
# Overall Mean Time For Marathon
average <- c(0.08837516, 0.07341435, 0.09006440, 0.09929210)
mean(average)

# Separating Men's Time
mens_data <- winners %>%
  filter(Category == "Men")

# Separating Women's Time
womens_data <- winners %>%
  filter(Category == "Women")

```


```{r}

# Plot Men and Women (Year vs Time)
ggplot() +
  geom_point(data = mens_data, aes(x = Year, y = Time), color = "blue") +
  geom_line(data = mens_data, aes(x = Year, y = Time), color = "blue", linetype = "solid") +
  geom_point(data = womens_data, aes(x = Year, y = Time), color = "pink") +
  geom_line(data = womens_data, aes(x = Year, y = Time), color = "pink", linetype = "solid") +
  labs(title = "London Marathon Winners' Times",
       x = "Year",
       y = "Time") +
  theme_minimal()

```

```{r}

# Create Table of Men vs Women Winning Times For Each Year

mvw_table <- winners %>%
  filter(Category %in% c("Men", "Women")) %>%
  group_by(Year) %>%
  summarize("Men's Mean" = mean(Time[Category == "Men"]), # Mean used to account for double winners
            "Women's Mean" = mean(Time[Category == "Women"]),
            "Annual Mean" = mean(Time[Category %in% c("Men", "Women")]))


```


```{r}
# Calculating Difference Between Men and Women's Times Each Year

time_difference_table <- winners %>%
  filter(Category %in% c("Men", "Women")) %>%
  group_by(Year, Category) %>%
  summarize(Mean_Time = mean(Time)) %>% # Mean Used to Account For Double Winner
  spread(Category, Mean_Time) %>%
  mutate(Difference = Women - Men) %>% # Women always have higher time,
  # This keeps difference positive / always says how much faster men were
  select(Year, Men, Women, Difference)

```

```{r}
# Plotting Differences Line Plot

ggplot(time_difference_table, aes(x = Year, y = Difference, group = 1)) +
  geom_line(color = "green", size = 1) +
  geom_point(color = "green", size = 3) +
  labs(title = "Difference Between Men and Women's Times (London Marathon)",
       x = "Year",
       y = "Time Difference") +
  theme_minimal()

# Using Regression Plot and Smooth to look at linearity and increasing/decreasing

ggplot(time_difference_table, aes(x = Year, y = Difference)) +
  geom_point(color = "green", size = 3, alpha = 0.7) +
  geom_smooth(method = "loess", color = "blue", se = FALSE) +  # Blue smooth line
  geom_smooth(method = "lm", se = FALSE, color = "red") +  # Red regression line
  labs(title = "Difference Between Men and Women's Times (London Marathon)",
       x = "Year",
       y = "Time Difference") +
  theme_minimal()
```

Prompts to interpret differences: 

on average, men tend to have faster winning times in the London Marathon compared to women. Consider discussing potential factors influencing this difference, such as training methods, athleticism, or other variables.

Training and Preparation: Discuss how training regimens and preparation strategies might differ between men and women. Consider whether advancements in coaching, nutrition, or sports science have contributed to changes in performance.

