---
title: "ST661 - Group Project - London Marathon"
author: 'Group 2: Sam Alexander White, Bimal Oommen John, Akshay Suresh Varma and
  Fernando Nicolas Szeliga Cabezuelo'
date: "2023-11-25"
output:
  pdf_document: default
  html_document:
    code_folding: hide
---

```{r, include=FALSE}

#Loading Packages

library(tidyverse)
library(GGally)
library(MASS)
library(mosaicData)
library(gridExtra)

```

```{r, message=FALSE}

#data downloading

remotes::install_github("nrennie/LondonMarathon")
data(winners, package = "LondonMarathon")
data(london_marathon, package = "LondonMarathon")

#Joint London Marathon and Winners by Year

joint_marathondata <- dplyr::full_join(london_marathon, winners, by="Year")

# Time conversion from fraction of days to h:m:s

# The time column is in fraction of days,so this was converted to seconds
time_in_seconds <- as.numeric(joint_marathondata$Time) * 24 * 60 * 60

# Formatting the time components
joint_marathondata$Time_Converted <- format(as.POSIXct(time_in_seconds,origin = "1970-01-01", tz = "UTC"), format = "%H:%M:%S")


# NEW VARIABLE: GEOGRAPHICAL LOCATION

winners_countries <- unique(joint_marathondata$Nationality)


geolocation_vector <- sapply(joint_marathondata$Nationality, function(x)
{
  
  if (x== "Norway" | x=="Ireland" |x== "Denmark" |x== "Poland" |x== "Sweden" | x=="Portugal" |x== "France" |x== "Germany" |x== "Belgium" |x== "Switzerland" |x== "Spain" |x== "Italy" | x=="Netherlands")
  {
    return("Rest of Europe")
  }
    else if (x== "United Kingdom")
  {
    return("United Kingdom")
  }
  else if (x== "United States" |x== "Canada" |x== "Mexico")
  {
    return("North America")
  }
  else if (x== "Japan" |x== "Soviet Union" |x== "China")
  {
    return("Asia")
  }
  else if (x== "Kenya" |x== "Morocco" |x== "Ethiopia")
  {
    return("Africa")
  }
  else (x == "Australia")
  {
    return("Oceania")
  }
  
})

joint_marathondata$`Winner's Geographical Location` <- geolocation_vector
    
```


### Introduction

#### Data Description

The data consists of information about the London Marathon over the years, including details about the number of applicants, accepted participants, starters, finishers, and the amount raised for charity. Additionally, there is a second table containing information about the winners in different categories (Men, Women, Wheelchair Men, Wheelchair Women) over the years.

Exploring the London Marathon data, we aim to uncover trends related to participant distribution, charity fundraising, and winners' performances across categories. Notably, we'll investigate if Kenya, with the most Male/Female winners, also secures the fastest victories. Additionally, we'll examine whether the UK, having the highest handicap winners, demonstrates similar performance trends in the Male/Female categories. Another focal point involves determining if men's winning times significantly differ from women's accomplishments.

The main dataset includes information on applicants, accepted participants, starters, finishers, and charity fundraising for each year. The second dataset provides details about the winners in different categories, including their names, nationalities, and finishing times. 

#### Data Overview


```{r, fig.cap= "..."}
ggplot(data = joint_marathondata) +
  geom_bar(mapping = aes(x = `Winner's Geographical Location`, fill=Category), position = "fill") + labs(x= "Geographic Area", y= "Density", title =  "Figure 1. Winner's Geographic Distribution by Competition") + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

```


The leading country in the Men and Female categories is Kenya, with 17 and 14 winners in each category respectively. Second in the male category is the United Kingdom (UK) with 6 winners, followed by Ethiopia with 5 winners. In the female category, Kenya is followed by the UK with 7 winners and Norway with 6 winners.

In the handicapped Male and Female categories, the country with the highest number of winners is the UK, with 16 winners in the male category and 15 in the female category. The UK is followed by Switzerland in the Men category with 6 winners and the United States (US) in the Female category with 6 winners as well. 

#### Research Questions

The following three research questions are being assessed in this report:

**Research Question 1: Are the Kenyan winners the fastest overall in the male and female competition?**

As mentioned in the data overview, the highest number of winners in the male a female categories are from Kenya. In addition, the number of winners from Kenya is 50% or higher than the number of winners from the UK, which is the host country. Hence, with the data available it is possible to analyse whether Kenyan runners are also the fastest overall and if that is a factor to make them the most successful country in these two categories. 


**Research Question 2: ?**


**Research Question 3: Are men's finish times similar to women's finish time?**



### Data Wrangling

We combined the "winners" and "London Marathon" tables by matching the "year" to create a single dataset for the London Marathon. Our main focus is on analyzing the winners to understand them better. To do this, we filtered the data to spotlight the winners and added a new date variable to fix formatting issues. Notably, we didn't collect information on the 2022–2023 handicap and found gaps in charity and money raised data. This ensures our analysis hones in on crucial aspects.

We handled variable types well, especially converting the "Time" variable from fraction-of-days to hours, seconds, and minutes for more meaningful insights into finish times. This ensures our data is in the right format for analysis.

There are some interesting points in the data. In 2020, there were many applicants, likely affected by COVID-19, leading to fewer accepted participants. Also, in 1991, a winner is listed from the Soviet Union, but remember, the Soviet Union dissolved that year. These quirks add intrigue to the data and highlight the importance of careful analysis.


In the inaugural London Marathon of 1981, 7,741 registered participants, with 6,255 completing the race. Dick Beardsley and Inge Simonsen made history by tying in the Men's Elite Race, while Joyce Smith won the Women's Elite Race. Wheelchair races started in 1983, and British athletes have clinched 40 victories.

British success continued with David Weir's 2012 Men's Wheelchair Race win. Kenya dominates with 17 Men and 14 Women winners. In handicapped categories, the UK leads with 16 Men and 15 Women winners, followed by Switzerland and the US.

Course records, including Eliud Kipchoge's Men's Elite Race (2:02:37) and Paula Radcliffe's Women's Elite Race (2:15:25), showcase remarkable performances. In 2021, Marcel Hug and Manuela Schär set new records in Wheelchair Races.

```{r}

```

### Analysis

##### **Research Q1** : are the Kenyan winners in the Male and Female categories the fastest? 

```{r,fig.cap= "..."}

# Time comparison by Nationality Plot. Category: Men

joint_marathondata |> filter(Category=="Men") |> group_by(Nationality) |> summarise(finish_time= mean(Time)) -> men_category


# The time column is in fraction of days,so this was converted to seconds
time_men <- as.numeric(men_category$finish_time) * 24 * 60 * 60

# Formatting the time components
men_category$time_conversion <- format(as.POSIXct(time_men,origin = "1970-01-01", tz = "UTC"), format = "%H:%M:%S")

p_male <- ggplot(men_category, aes(x = reorder(Nationality,finish_time, decreasing=TRUE), y = time_conversion)) +  geom_point() + geom_segment(aes(xend = fct_reorder(Nationality,finish_time)), yend = 0, col = "blue") +
  coord_flip() + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) + labs(x= "Nationality", y= "Time", title =  "Men Mean Finish Time")


# Time comparison by Nationality Plot. Category: Women

joint_marathondata |> filter(Category=="Women") |> group_by(Nationality) |> summarise(finish_time= mean(Time)) -> women_category


# The time column is in fraction of days,so this was converted to seconds
time_women <- as.numeric(women_category$finish_time) * 24 * 60 * 60

# Formatting the time components
women_category$time_conversion <- format(as.POSIXct(time_women,origin = "1970-01-01", tz = "UTC"), format = "%H:%M:%S")

p_female <- ggplot(women_category, aes(x = reorder(Nationality,finish_time, decreasing=TRUE), y = time_conversion)) +
  geom_point() + geom_segment(aes(xend = fct_reorder(Nationality,finish_time)), yend = 0, col = "purple") +
  coord_flip() + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) + labs(x= "Nationality",y= "Time", title =  "Women Mean Finish Time")


grid.arrange(p_male, p_female, ncol = 2)


```




### Conclusions
